---
title: "ME 317: Statistical Methods in Risk Management at LSE"
author: Andrea Ranzato;
date: August 1st, 2019
output: 
  github_document:
     pandoc_args: --webtex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE,
                      fig.align = "center", fig.width=1.41*6, fig.height=1*6, 
                      fig.margin=TRUE)
```
```{r packages}
library(tidyverse)
library(tidyquant)
library(lubridate)
library(broom)
library(scales)
library(knitr)
library(broom)
library(qqplotr)
library(scales)

colours <- c("#7cb5ec", "#434348", "#90ed7d", "#f7a35c", "#8085e9", 
             "#f15c80", "#e4d354", "#2b908f", "#f45b5b", "#91e8e1")
```
```{r dataraw}
dataraw <- read_csv("../data/stock-prices.csv") %>% 
  select(-c("PERMNO","VOL")) %>% 
  mutate(DATE = ymd(date),
         TICKER = factor(TICKER)) %>% 
  select(-date) %>% 
  relocate(DATE, .before = everything()) %>% 
  mutate(Sector = case_when(
    TICKER %in% c("AIG","BA","GS","MS","NTRS","WFC") ~ "Financial",
    TICKER %in% c("AMZN","KO") ~ "Consumer",
    TICKER == "XOM" ~ "Energy",
    TICKER == "MSFT" ~ "IT")) %>% 
  mutate(Sector = factor(Sector))
```

## Summary

The general purpose of this project is to use techniques learned during the course in *Statistical methods in risk management* useful to analyze stock price data. In particular, we consider prices concerning ten different stocks recorded during the 2007-2009 financial crisis. The data source is the *Wharton Research Data Services*. 

The analysis is structured as follows. In the first section, I show how two extreme events, the rescue of Bear Sterns on March 14, 2008 and Lehman Brothers bankruptcy on September 15, 2008, affected the price and log-return dynamics. Next, I test both the marginal and joint normality of the ten stock returns by means of the *Jarque-Bera* and *Mardia test* respectively. First, we observe that the normal distribution fails to capture returns happening in the tails. Second, joint normality is not detected. Subsequently, I evaluate some pair-wise stock returns, either from companies of the same sector and different ones, in order to quantify their correlation. 

In the second section, I focus on financial stocks with the objective of assessing their dependence using copulas of different types. We see, that the copula which fits the observations best is the t-Copula.

In the third section, I consider a linear portfolio which invest \$1000 in each of the stocks at the beginning of the period, assuming that its composition does not change during the entire period. Afterwards, I evaluate the Value-at-Risk of the portfolio at 95% level using two different strategies. Firstly, I compute the VaR using the empirical profit and loss distribution, secondly using a normal model. The comparison between the two shows that the parametric approach would have led to more conservative capital management decisions compared to empirical one, since it signals higher potential losses. In the end, I add to the analysis 2010 portfolio returns, in order to record the number of violations occurred under the two different frameworks. We see that the normal model incurs in fewer violations with respect to the the historical one.

The last section is devoted to the description of the topics covered during the course. 

## Prices and log returns dynamics

The purpose of this section consists in showing some empirical properties of financial time series using a sample of US assets.
First of all, I conduct an exploratory analysis of the price dynamics of equities in the Table below between 3 January 2007 and 31 December 2009. 
As we can see in the Figure below, a strong negative trend plays out in correspondence with Bear Stearns rescue on March 14 2008 and terminates around January 20 2009. Table~\ref{tab:percentagechange} quantifies the percentage price change recorded in that period summarizing the worst downturn of the US financial system since the Great Depression. 

```{r stocks}
stocks <- tibble(Company = c("American International Group","Amazon.com, Inc.",
                             "Bank of America Corp.","Goldman Sachs Group, Inc.",
                             "Coca-Cola Company","Morgan Stanley",
                             "Microsoft Corp.","Northern Trust Corp.",
                             "Wells Fargo & Co","ExxonMobil Corp."),
                 Sector = c("Financials","Consumer","Financials","Financials",
                            "Consumer","Financials","IT","Financials","Financials",
                            "Energy"),
                 Ticker = c("AIG","AMZN","BA","GS","KO","MS",
                            "MSFT","NTRS","WFC","XOM"))
kable(stocks, caption = "Table: Equity portfolio composition.",)
```
```{r percentage-price-change}
dataraw %>% 
  filter(DATE %in% c(ymd("2008-03-14"), ymd("2009-01-20"))) %>% 
  pivot_wider(names_from = DATE, values_from = PRC) %>% 
  rename(P2 = "2009-01-20") %>%
  rename(P1 = "2008-03-14") %>%
  mutate(DELTA = ((P2-P1)/P1)*100) %>% 
  kable(caption = "Table: Percentage price change between 14 Mar. 2008 and 20 Jan. 2009",
        col.names = c("Ticker", "Sector", "P.2008-03-14", "P.2009-01-20", "(%) Change"))
```

```{r percentage-price-change-graph}
dataraw %>% 
  filter(DATE %in% c(ymd("2008-03-14"), ymd("2009-01-20"))) %>% 
  arrange(DATE) %>% 
  mutate(DATE = factor(DATE)) %>% 
  ggplot(aes(DATE, PRC, group = TICKER)) +
  geom_line(aes(color = Sector), size = 1.05) +
  geom_point(aes(color = Sector), size = 1.25) +
  scale_fill_brewer() +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "top") +
  labs(y = "Price ($)",
       x = "Days",
       title = "Change in Prices from March 14, 2008 to January 20, 2009") 
```

```{r price-dynamics}
ggplot(dataraw, aes(x = DATE, y = PRC, group = TICKER, colour = TICKER)) + 
  geom_line(size = 0.5) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "top") +
  # Lehman bankruptcy
  geom_vline(xintercept = ymd("2008-09-15"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  # Bear Stearns rescue
  geom_vline(xintercept = ymd("2008-03-14"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  labs(title = "Stock Price Dynamics",
       subtitle = "January 03, 2007 - December 31, 2009\nBear Stearns rescue; Lehman's bankruptcy",
       caption = "Source: Wharton Research Data Services",
       x = "Date",
       y = "Price") +
  scale_colour_manual(values = colours)
```

```{r log-returns}
logReturns <- dataraw %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC) %>% 
  mutate(across(where(is.numeric), ~ log(.x))) %>% 
  mutate(across(where(is.numeric), ~ .x - lag(.x, n = 1))) %>% 
  slice(-1) 
```

```{r log-returns-graph, fig.width=1.41*6,fig.height=8}
logReturns %>% 
  pivot_longer(cols = !DATE, names_to = "TICKER", values_to = "LOGPRICE") %>% 
  arrange(TICKER) %>% 
  ggplot(aes(x = DATE, y = LOGPRICE, colour = TICKER)) +
  geom_line(size = 0.35) + 
  facet_wrap(~ TICKER, ncol = 2, nrow = 5) +
  theme_light() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  scale_y_continuous(limits = c(-1,1)) +
  scale_colour_manual(values = colours, guide = FALSE) +
  # Lehman bankruptcy
  geom_vline(xintercept = ymd("2008-09-15"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  # Bear Stearns rescue
  geom_vline(xintercept = ymd("2008-03-14"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  labs(y = "Log Prices",
       x = "Date",
       title = "Stock Performance Based on Log Returns",
       subtitle = "January 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services")

```

```{r jarque-bera-test}
logReturns %>% 
  pivot_longer(cols = !DATE, names_to = "TICKER", values_to = "PRICE") %>% 
  select(-DATE) %>% 
  nest(ts = c(PRICE)) %>% 
  mutate(ts = map(ts, ~ as.matrix(.x))) %>% 
  mutate(ts = map(ts, ~ tseries::jarque.bera.test(.x))) %>% 
  mutate(ts = map(ts, tidy)) %>% 
  unnest(ts) %>% 
  mutate(Normal = "No") %>% 
  select(-c(parameter,method)) %>% 
  arrange(statistic) %>% 
  kable(caption = "Table: Jarque Bera Test, n. obs. = 756",
        digits = 2,
        format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE))
```

```{r marginal-normality-QQplot, fig.show="hold", results='hide', fig.height=18}
logReturnsList <- logReturns  %>% 
  select(!DATE) %>% 
  pivot_longer(cols = everything(), names_to = "TICKER", values_to = "PRC") %>% 
  group_by(TICKER) %>% 
  nest() %>% 
  mutate(data = map(data, ~ as.data.frame(.x))) %>% 
  pull(data) %>% 
  set_names(nm = unique(dataraw$TICKER))

tickers <- names(logReturnsList)

qqPlots <- vector(mode = "list", length = length(logReturnsList))
for (i in seq_along(logReturnsList)) {
  qqPlots[[i]] <- ggplot(logReturnsList[[i]], aes(sample = PRC)) +
    qqplotr::stat_qq_line(colour = "black", size = 0.7) +
    qqplotr::stat_qq_point(size = 0.6, colour = "#696969") +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
         title = str_c("Stock: ", tickers[i])) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
}

gridExtra::grid.arrange(grobs = qqPlots, nrow = 5)
```
```{r student-t-test}
# One-sample Kolmogorov-Smirnov test for student-t distribution
logReturns %>% 
  select(!DATE) %>% 
  as.list() %>% 
  map(~ ks.test(.x, "pt", df = 10)) %>% 
  enframe() %>% 
  mutate(value = map(value, ~ tidy(.x))) %>% 
  unnest(value) %>% 
  select(-method) %>% 
  arrange(statistic) %>% 
  mutate(t.Student = rep("NO", 10)) %>% 
  kable(caption = "Table: One-sample Kolmogorov-Smirnov test for student-t distribution, n. obs. = 756, dgf. = 10",
        col.names = c("Ticker", "Statistic","p-value", "Alternative", "t-Student"),
        digits = 4,
        format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE))
```

```{r marginal-t-student-QQplot, fig.show="hold", results='hide', fig.height=18}
qqPlots2 <- vector(mode = "list", length = length(logReturnsList))
for (i in seq_along(logReturnsList)) {
  qqPlots2[[i]] <- ggplot(logReturnsList[[i]], aes(sample = PRC)) +
    qqplotr::stat_qq_line(distribution = "t", dparams = 10, colour = "black", size = 0.7) +
    qqplotr::stat_qq_point(size = 0.6, colour = "#696969") +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
         title = str_c("Stock: ", tickers[i])) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5))
}
gridExtra::grid.arrange(grobs = qqPlots2, nrow = 5)
```

```{r log-returns-kurtosis-skewness}
dataraw %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC) %>% 
  mutate(across(where(is.numeric), ~ log(.x))) %>% 
  mutate(across(where(is.numeric), ~ .x - lag(.x, n = 1))) %>% 
  slice(-1) %>% 
  select(!DATE) %>% 
  MVN::mvn(mvnTest = "mardia") %>% 
  pluck(3) %>% 
  select(Skew, Kurtosis) %>% 
  arrange(Kurtosis) %>% 
  kable(caption = "Table: Log Returns Skewness and Kurtosis, n. obs. = 756",
        col.names = c("Skewness", "Kurtosis"),
        digits = 2,
        format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE))
```

```{r multivariate-normality}
normality_test <- dataraw %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC) %>% 
  mutate(across(where(is.numeric), ~ log(.x))) %>% 
  mutate(across(where(is.numeric), ~ .x - lag(.x, n = 1))) %>% 
  slice(-1) %>% 
  select(!DATE) %>% 
  MVN::mvn(mvnTest = "mardia")

normality_test$multivariateNormality %>% 
  slice(1:2) %>% 
  mutate(Statistic = as.double(levels(Statistic))) %>% 
  kable(caption = "Table: Mardia test for Multivariate Normality.", digits = 3,
        format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE))
```

```{r multivariate-normality-graph}
normality_test <- dataraw %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC) %>% 
  mutate(across(where(is.numeric), ~ log(.x))) %>% 
  mutate(across(where(is.numeric), ~ .x - lag(.x, n = 1))) %>% 
  slice(-1) %>% 
  select(!DATE) %>% 
  MVN::mvn(mvnTest = "mardia", multivariatePlot = "qq")
```

## Copulas

### Multivariate Copulas

```{r financial-stocks}
finStocks <- dataraw %>% 
  filter(Sector == "Financial") %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = "TICKER", values_from = "PRC") %>% 
  mutate(across(where(is.numeric), ~ log(.x))) %>% 
  mutate(across(where(is.numeric), ~ .x - lag(.x, n = 1))) %>% 
  slice(-1) %>% 
  select(-DATE) %>% 
  as.matrix()
```

```{r pseudo-observations, fig.dim=c(10,10)}
#library(QRM)
library(copula)
pseudos <- copula::pobs(finStocks)

# Equivalently to find pseudo-observations
# as_tibble(finStocks) %>% 
#   mutate(across(everything(), ~ QRM::edf(.x, adjust = 1)))
```

```{r fit-multivariate-normal-copula}
normalCopula <- fitCopula(normalCopula(dim = ncol(finStocks), dispstr = "un"), 
                           data = pseudos, 
                           method = "mpl")
logLikNormal <- summary(normalCopula)$loglik
```

```{r fit_multivariate-student-t}
tStCopula <- fitCopula(tCopula(dim = ncol(finStocks), dispstr = "un"),
                       data = pseudos,
                       method  = "mpl")
logLikStudent <- summary(tStCopula)$loglik
```

```{r fit-multivariate-gumbel}
gumbCompula <- fitCopula(gumbelCopula(dim = ncol(finStocks)), 
                                      data = pseudos,
                         method = "mpl")
logLikGumbel <- summary(gumbCompula)$loglik
```

```{r fit-multivariate-clayton}
clayCompula <- fitCopula(gumbelCopula(dim = ncol(finStocks)), 
                                      data = pseudos,
                         method = "mpl")
logLikClayton <- summary(clayCompula)$loglik
```

### Bivariate Copulas

```{r pairwise-scatter-financials, fig.dim=c(10,10)}
logReturns %>% 
  select(c("AIG","BA","GS","MS","NTRS","WFC")) %>%
  GGally::ggpairs(lower = list(continuous = GGally::wrap("points", size = 0.1))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        strip.text = element_text(face="bold", size=rel(0.60)),
        plot.title=element_text(size = rel(1.1), 
                                lineheight = .9, 
                                family = "Helvetica",
                                face = "plain", 
                                colour = "black",
                                hjust = 0.5)) +
  labs(title = "Pair-wise Returns scatter plots of Financial Stocks")
```

```{r pairwise-financial-mardia-test}
combs <- t(combn(1:ncol(finStocks), m = 2))

combsName <- c()
values <- vector(mode="list", length = nrow(combs))

for (i in 1:nrow(combs)) {
  combsName[i] <- str_c(colnames(finStocks[,combs[i, ]]), collapse = "-")
  values[[i]] <- MVN::mvn(finStocks[,combs[i, ]], 
                          mvnTest = "mardia")$multivariateNormality
}

values %>% 
  set_names(nm = combsName) %>% 
  enframe() %>% 
  mutate(value = map(value, ~ as_tibble(.x))) %>% 
  unnest(value) %>% 
  filter(Test != "MVN") %>% 
  mutate(across(where(is.factor), ~ as.character(.x))) %>% 
  mutate(Statistic = as.numeric(Statistic)) %>% 
  rename(P.value = "p value") %>% 
  mutate(P.value = as.numeric(P.value)) %>% 
  group_by(name) %>% 
  nest() %>% 
  ungroup() %>% 
  mutate(data = map2(data, combsName, 
                   ~ kable(.x, caption = str_c("Normality test between: ", .y),
                           digits = 2, 
                           format.args = list(big.mark = ".", 
                                              decimal.mark = ",", 
                                              scientific = FALSE)))) %>% 
  pull(data) %>% 
  kables()
```

```{r empirical-scatter-pseudo-observations, fig.dim=c(8,8), fig.cap="Figure (a): Pseudo observations (uniform margins) obtained from actual data, with non parametric estimation of the CDFs."}
splom2(pseudos, cex = 0.2, col.mat = "black", pch = 16)
```

```{r fit-bivariate-t-copula}
bivStudCopula <- vector(mode = "list", length = nrow(combs))
for (i in 1:nrow(combs)) {
  bivStudCopula[[i]] <- fitCopula(tCopula(dim = 2, dispstr = "un"), 
                                  data = pseudos[, combs[i,]],
                                  method = "mpl")
}
names(bivStudCopula) <- combsName
```

```{r multivariate-student-t-simulation, fig.dim=c(8,8), fg.cap="Figure (b): 755 simulated pseudo-observations obtained from a multivariate Student-t distribution with 4 dof and equicorrelation dispersion matrix with rho = 0.7"}
nObs <- nrow(finStocks)
tNVar <- ncol(finStocks)
cor.par <- 0.7

# Dispersion matrix
tSigma <- (1-cor.par)*diag(tNVar) + matrix(cor.par, c(tNVar,tNVar), nrow = 6)
t.dof <- 4

# Multivariate dim 6 student t distribution (margins are uniform)
tData <- mnormt::rmt(nObs, mean = rep(0, tNVar), S = tSigma, df = t.dof)
tCopulaData <- apply(tData, 2, pt, df = t.dof)
splom2(tCopulaData, cex = 0.2, col.mat = "black", pch = 16)

# Contour charts
distribution.limits <- c(0.01, 0.99)
Sigma.contour <- QRM::equicorr(2, 0.5)
threed.limits <- c(-3, 3)

par(mfrow = c(1,2))
QRM::BiDensPlot(func = QRM::dcopula.t, xpts = distribution.limits, ypts = distribution.limits,
           df = t.dof, Sigma = Sigma.contour, type = "contour")

QRM::BiDensPlot(func = QRM::dcopula.t, xpts = distribution.limits , 
                ypts = distribution.limits, df = t.dof, Sigma=Sigma.contour, 
                type = "persp", npts =15)

```

```{r summary-log-likelihood}
loglikelihoods <- tibble(logLikNormal,logLikStudent,logLikGumbel,logLikClayton)
kable(loglikelihoods, 
      caption = "Table: MLE results of four different multivariate copula of the Financial stocks",
      digits = 2, 
      format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE),
      col.names = c("Normal","Student","Gumbel","Clayton"))
```

```{r histograms-pseudo-obs, results="hide", fig.height=8, fig.cap="Figure: Pseudo-observations Distribution"}
finNames <- sort(c("AIG","BA","GS","MS","NTRS","WFC"))

uniHist <- pobs(finStocks) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything()) %>% 
  arrange(name) %>% 
  nest(value) %>% 
  #pull(data) %>% 
  mutate(plots = map2(data, finNames, ~ ggplot(.x, aes(x = value)) + 
                        geom_histogram(colour = "grey", 
                                       fill = "lightblue",
                                       bins = 30) +
                        theme_light() +
                        theme(plot.title = element_text(hjust = 0.5)) +
                        labs(title = str_c("Stock: ", .y)))) %>% 
  pull(plots)
gridExtra::grid.arrange(grobs = uniHist, ncol = 2)
```

```{r fit-bivariate-normal}
bivNormalCopula <- vector(mode="list", length = nrow(combs))
for (i in 1:nrow(combs)) {
  bivNormalCopula[[i]] <- fitCopula(normalCopula(dim = 2), 
                                    data = pseudos[,combs[i,]],
                                    method = "mpl")
}
names(bivNormalCopula) <- combsName
```

```{r fit-bivariate-gumbel}
bivGumbelCopula <- vector(mode = "list", length = nrow(combs))
for (i in 1:nrow(combs)) {
  bivGumbelCopula[[i]] <- fitCopula(normalCopula(dim = 2), 
                                    data = pseudos[,combs[i,]],
                                    method = "mpl")
}
names(bivGumbelCopula) <- combsName
```

```{r fit-bivariate-clayton, eval=FALSE}
# bivClaytonCopula <- vector(mode = "list", length = nrow(combs))
# for (i in 1:nrow(combs)) {
#   bivClaytonCopula[[i]] <- fitCopula(claytonCopula(dim = 2),
#                                      data = pseudos[,combs[i,]],
#                                      method = "mpl")
# }
# names(bivClaytonCopula) <- combsName
```

## Portfolio Value-at-Risk

```{r portfolio-value}
# The value of the portfolio at the beginning of the period is $10.000
prices <- dataraw %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC) %>% 
  select(-DATE)

prices0 <- slice(prices, 1) %>% 
  select(everything()) %>% 
  unlist()

capital0 <- rep(1000, ncol(prices))
names(capital0) <- colnames(prices)
pfValue0 <- sum(capital0)

# Number of stocks purchased at time t=0 is kept fixed along successive time 
# periods
nAssets0 <- capital0/prices0
names(nAssets0) <- str_c("n.", colnames(prices))

# Portfolio value
PV <- as.matrix(prices)%*%as.matrix(nAssets0, nrow = nrow(prices))
PV %>% 
  as_tibble_col(column_name = "PV") %>% 
  mutate(DATE = unique(dataraw$DATE)) %>% 
  relocate(DATE, .before = everything()) %>% 
  ggplot(aes(DATE, PV)) +
  geom_line() +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
    # Lehman bankruptcy
  geom_vline(xintercept = ymd("2008-09-15"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  # Bear Stearns rescue
  geom_vline(xintercept = ymd("2008-03-14"), colour = "#696969", 
             size = 0.5, linetype = 2) +
  labs(title = "Portfolio value",
       subtitle = "Inital number of stocks kept fixed for the entire time framework.\nJanuary 03, 2007 - December 31, 2009",
       x = "Date",
       y = "($)")
```
```{r portfolio-weights}
# Find the weights of each stock at time t by dividinding its values by the 
# value of the portfolio at time t
PV <- PV %>% 
  as_tibble() %>% 
  mutate(DATE = unique(dataraw$DATE)) %>% 
  relocate(DATE, before = everything())

prices <- prices %>% 
  mutate(DATE = unique(dataraw$DATE)) %>% 
  relocate(DATE, before = everything()) 

nAssets0 <- as_tibble(nAssets0)

left_join(prices, PV) %>% 
  rename(PV = V1) %>% 
  relocate(PV, .after = DATE) %>% 
  pivot_longer(cols = !c(DATE,PV), names_to = "TICKER", values_to = "PRC") %>% 
  mutate(n = unlist(rep(nAssets0, nrow(prices)))) %>% 
  group_by(TICKER) %>% 
  nest() %>% 
  mutate(data = map(data, ~ mutate(.x, w = (PRC*n)/PV))) %>% 
  unnest() %>% 
  select(TICKER, DATE, w) %>% 
  ggplot(aes(x = DATE, y = w, colour = TICKER)) +
  geom_line() +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "top") +
  labs(title = "Portfolio Weights Dynamics",
       subtitle = "Inital number of stocks kept fixed for the entire time framework.\nJanuary 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services",
       x = "Date",
       y = "Weight") +
  scale_colour_manual(values = colours)

```

```{r portfolio-profit-loss, fig.height=10}
gg1 <- PV %>% 
  mutate(PL = -(V1 - lag(V1, n = 1))) %>% 
  ggplot(aes(x = DATE, y = PL)) +
  geom_line(size = 0.35) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  # Lehman bankruptcy
  geom_vline(xintercept = ymd("2008-09-15"), colour = "red", 
             size = 0.5, linetype = 2) +
  # Bear Stearns rescue
  geom_vline(xintercept = ymd("2008-03-14"), colour = "red", 
             size = 0.5, linetype = 2) +
  labs(title = "Absolute Portfolio Profit and Loss",
       subtitle = "January 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services",
       y = "p&l",
       x = "Date") 

gg2 <- PV %>% 
  mutate(PL = -(V1 - lag(V1, n = 1))) %>% 
  ggplot(aes(x = PL)) +
  geom_histogram(fill = "lightblue", colour = "grey", bins = 40) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  labs(title = "Absolute Portfolio Profit and Loss",
       subtitle = "January 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services",
       x = "p&l") 

gg3 <- PV %>% 
  mutate(PL = -(V1 - lag(V1, n = 1))) %>% 
  ggplot(aes(x = PL)) +
  geom_boxplot() +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  labs(title = "Absolute Portfolio Profit and Loss",
       subtitle = "January 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services",
       x = "p&l") 

plPlots <- list(gg1, gg2, gg3)

gridExtra::grid.arrange(grobs = plPlots, ncol = 1)
```

```{r VAR}
PL <- PV %>% 
  mutate(PL = -((V1 - lag(V1, n = 1))/V1)) %>% 
  pull(PL)
  
meanPL <- PV %>% 
  mutate(PL = (V1 - lag(V1, n = 1))/V1) %>% 
  summarise(mean = mean(PL, na.rm = TRUE)) %>% 
  pull()

sdPL <- PV %>% 
  mutate(PL = (V1 - lag(V1, n = 1))/V1) %>% 
  summarise(sd = sd(PL, na.rm = TRUE)) %>% 
  pull()

# Var as a percentage of portfolio’s current value
varGauss <- qnorm(0.95)*sdPL - meanPL
names(varGauss) <- "95%"
varHistorical <- quantile(PL, p = 0.95, na.rm = TRUE)

kable(matrix(c(varGauss,varHistorical), ncol=2), 
      col.names = c("Gaussian VaR", "Historical VaR"),
      digits = 4, 
      format.args = list(big.mark = ".", decimal.mark = ",", scientific = FALSE),
      caption = "95% Value-at-Risk calculated on 31 December 209 for a one-day horizon under parametric and non-parametric appproach.")

percPL <- PV %>% 
  mutate(percPL = -((V1 - lag(V1, n = 1))/V1)) %>% 
  slice(-1)

meanPercPl <- percPL %>% 
  summarise(mean = mean(percPL)) %>% 
  pull()

sdPercPl <- percPL %>% 
  summarise(sd = sd(percPL)) %>% 
  pull()

densSupport <- density(percPL$percPL)
yKernel <- approx(densSupport$x, densSupport$y, xout = varHistorical)$y

segmentsCoord <- data.frame(x1 = varHistorical, y1 = yKernel,
                            x2 = varGauss, y2 = dnorm(varGauss, 
                                                      mean = meanPercPl,
                                                      sd = sdPercPl))
percPL %>% 
  ggplot(aes(x = percPL)) +
  geom_density(colour = "black") +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  ggplot2::stat_function(fun = dnorm, 
                         args = list(mean = meanPercPl, sd = sdPercPl),
                         colour = "red") +
  geom_segment(aes(x = x1, xend = x1, y = 0, yend = y1), data = segmentsCoord) +
  geom_segment(aes(x = x2, xend = x2, y = 0, yend = y2), 
               colour = 'red', data = segmentsCoord) +
  labs(title = "Historical and Normal Value-at-Risk (one day horizon)",
       subtitle = "January 03, 2007 - December 31, 2009",
       caption = "Source: Wharton Research Data Services",
       x = "p&l")
```

```{r var-backtesting}
# Contains the 2010 financial year
varBacktesting <- read_csv("../data/var-backtesting.csv")

entireSerie <- bind_rows(dataraw, varBacktesting) %>% 
  select(-Sector) %>% 
  pivot_wider(names_from = TICKER, values_from = PRC)
  
PVentire <- as.matrix(entireSerie[,-1])%*%as.matrix(nAssets0, 
                                                    nrow = ncol(entireSerie))

PLentire <- PVentire %>% 
  as_tibble() %>% 
  mutate(absPL = -((value - lag(value, n = 1))/value)) %>% 
  mutate(DATE = entireSerie$DATE) %>% 
  relocate(DATE, before = everything())

valueAtRiskGauss <- function(vec, alpha) {
  muHat = mean(vec, na.rm = TRUE)
  sdHat = sd(vec, na.rm = TRUE)
  VaR = muHat + sdHat * qnorm(alpha)
  return(VaR)
}

valueAtRiskHistorical <- function(vec, alpha) {
  VaR = quantile(vec, alpha, na.rm = TRUE)
  return(VaR)
}

rollwindowSize <- (nrow(dataraw)/nrow(nAssets0))

# First computed VaR is on December 31, 2009 (one day horizon).
# We shift forward by one day the series because we are calculating the VaR
# over one-day risk horizon. In fact, "rollapply" locates the first result
# in correspondence of the time instant from which we account for the risk
# horizon. This operation allows to make a comparison between the realized
# loss and the one-day VaR.
#### Rolling Gaussian Value at Risk
rollGauss <- PLentire %>% 
  select(-value) %>% 
  tidyquant::tq_mutate(mutate_fun = rollapply,
                       align = "right",
                       width = rollwindowSize,
                       FUN = valueAtRiskGauss,
                       alpha = 0.95,
                       by.column = FALSE) %>% 
  filter(!is.na(value)) %>% 
  # Move ahead in order to compare actual loss with VaR
  mutate(value = lag(value, n = 1)) %>% 
  mutate(Violation = if_else(absPL > value, "YES", "NO"))

numberViolationG <- rollGauss %>% 
  slice(-1) %>% 
  count(Violation)

#### Rolling Hstorical Value at Risk
rollHist <- PLentire %>% 
  select(-value) %>% 
  tidyquant::tq_mutate(mutate_fun = rollapply,
                       align = "right",
                       width = rollwindowSize,
                       FUN = valueAtRiskHistorical,
                       alpha = 0.95,
                       by.column = FALSE) %>% 
  filter(!is.na(value)) %>% 
  # Move ahead in order to compare actual loss with VaR
  mutate(value = lag(value, n = 1)) %>% 
  mutate(Violation = if_else(absPL > value, "YES", "NO"))

numberViolationH <- rollHist %>% 
  slice(-1) %>% 
  count(Violation)

# Graph violations
viol1 <- rollGauss %>% 
  slice(-1) %>% 
  mutate(absPL = absPL) %>% 
  ggplot(aes(x = DATE)) +
  geom_line(aes(y = absPL)) +
  geom_line(aes(y = value, colour = "red")) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "none") +
  labs(subtitle = "95% one-day horizon Gaussian Value-at-Risk of the portfolio",
       title = "VaR backtest",
       caption = "Source: Wharton Research Data Services",
       x = "Date",
       y = "p&l")

viol2 <- rollHist %>% 
  slice(-1) %>% 
  mutate(absPL = absPL) %>% 
  ggplot(aes(x = DATE)) +
  geom_line(aes(y = absPL)) +
  geom_line(aes(y = value, colour = "red")) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "none") +
  labs(subtitle = "95% one-day horizon Historical Value-at-Risk of the portfolio",
       title = "VaR backtest",
       caption = "Source: Wharton Research Data Services",
       x = "Date",
       y = "p&l")

violations <- list(numberViolationG, numberViolationH) %>% 
  map2(c("Gaussian VaR", "Historical VaR"),
       ~ kable(.x, 
               caption = str_c("Number violations for 2010 returns under ", .y)))
kables(violations)

violationsPlots <- list(viol1, viol2)

gridExtra::grid.arrange(grobs = violationsPlots, ncol = 1)
```
